---
title: "Lab 10: Difference-in-differences"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Learning objectives

By the end of this lab, you should be able to:

1. Describe the 2×2 Difference-in-Differences (DiD) estimand and the Average Treatment Effect on the Treated (ATT) in this framework.
2. Compute a DiD estimate from group averages and compare it to the coefficient from a regression with an interaction term.
3. Explain why the parallel trends and no-anticipation assumptions matter, and how to think about them in an applied case.
4. Use the Lost-Stats 2×2 DiD example on marijuana legalization and murder rates to connect theory and practice.
5. Propose feasible extensions of a 2×2 DiD design for class projects.

# Why this lab?

This lab uses a clean 2×2 DiD example from the Lost-Stats classroom page. The example compares murder rates in states that legalized marijuana in 2014 to states that did not, before and after legalization.

That makes the design easy to understand because it has:

- two groups (treated vs control),
- two periods (pre-2014 vs post-2014), and
- one treatment event that happens at a single time.

We use this example to build intuition for:

- a cell-means DiD estimate,
- the equivalent regression formulation, and
- why the assumptions of **no anticipation** and **parallel trends** are central.

Key points:

- The DiD estimator compares the change in the treated group with the change in the control group.
- The treated group’s counterfactual is not the observed control mean after treatment; it is the treated group’s pre-post change if it had followed the control trend.
- The regression formulation uses `treat`, `after`, and their interaction.
- The main assumptions are:
  - **Parallel trends**: treated and control groups would have followed the same trend absent treatment.
  - **No anticipation**: treated units do not start changing outcomes before treatment begins.

# 1. Show data

## Packages

```{r packages}
install_if_missing <- function(pkgs) {
  missing <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
  if (length(missing) > 0) install.packages(missing)
}
install_if_missing(c("dplyr", "ggplot2", "broom", "knitr", "readxl", "httr"))

library(dplyr)
library(ggplot2)
library(broom)
library(knitr)
library(readxl)
library(httr)

theme_set(theme_minimal())
```

## Load the dataset

```{r load-data}
# Lost-Stats 2x2 DiD example with marijuana legalization and murder rates.
url <- "https://raw.githubusercontent.com/LOST-STATS/LOST-STATS.github.io/master/Model_Estimation/Data/Two_by_Two_Difference_in_Difference/did_crime.xlsx"

temp <- tempfile(fileext = ".xlsx")
GET(url, write_disk(temp, overwrite = TRUE))

DiD <- read_excel(temp)

DiD <- DiD %>%
  mutate(
    after = year >= 2014,
    treatafter = treat * after
  )

DiD
```

## Dataset overview

```{r data-overview}
DiD %>%
  summarise(
    n_obs = n(),
    n_treated = sum(treat == 1),
    n_control = sum(treat == 0),
    years = paste(min(year), max(year), sep = "-")
  ) %>%
  kable()
```

```{r data-glimpse}
glimpse(DiD)
```

> **Discussion:** The data are already aggregated at the treated-group level. What are the advantages and disadvantages of working with this collapsed dataset instead of individual-state data?

# 2. Do Diff-in-diff with averages

## Average outcomes by group and period

```{r mean-table}
means <- DiD %>%
  group_by(treat, after) %>%
  summarise(
    mean_murder = mean(murder, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )

means %>%
  mutate(group = case_when(
    treat == 1 & after == 0 ~ "Treated pre",
    treat == 1 & after == 1 ~ "Treated post",
    treat == 0 & after == 0 ~ "Control pre",
    treat == 0 & after == 1 ~ "Control post",
    TRUE ~ NA_character_
  )) %>%
  select(group, mean_murder, n) %>%
  kable(digits = 3)
```

## Hand compute the 2×2 DID estimate

```{r did-averages}
cell <- means %>%
  select(treat, after, mean_murder)

mu_T_pre  <- cell$mean_murder[cell$treat == 1 & cell$after == 0]
mu_T_post <- cell$mean_murder[cell$treat == 1 & cell$after == 1]
mu_C_pre  <- cell$mean_murder[cell$treat == 0 & cell$after == 0]
mu_C_post <- cell$mean_murder[cell$treat == 0 & cell$after == 1]

att_avg <- (mu_T_post - mu_T_pre) - (mu_C_post - mu_C_pre) # 
att_avg
```

> **Exercise 1:** Interpret the sign of the DID estimate. Does the result suggest that legalization increased or decreased murder rates in treated states relative to control states?

# 3. Then do with regression

## Regression specification

```{r did-regression}
mod_did <- lm(murder ~ treat * after, data = DiD)
summary(mod_did)

tidy(mod_did) %>%
  filter(term == "treat:afterTRUE")
```

## Regression with a group indicator only

```{r did-regression-group}
mod_group <- lm(murder ~ treat + after + treatafter, data = DiD)
summary(mod_group)
```

> **Discussion:** Why does the interaction term capture the policy effect while the main effects do not?

# 4. Then evaluate no-anticipation

## What no-anticipation means here

No anticipation means that treated states did not change murder rates in expectation of legalization before 2014. In this example, the policy should only affect the post period.

## Discussing the assumption

- What institutional facts would support or undermine no anticipation in this setting?
- If legalization was widely anticipated, how would that show up in the pre-period?

## Simulation-style thought experiment

```{r no-anticipation-thought}
# This is a conceptual exercise rather than a formal test.
# Suppose we pretend the treatment started in 2012 instead of 2014.
DiD_placebo <- DiD %>%
  mutate(after_placebo = year >= 2012,
         treatafter_placebo = treat * after_placebo)

mod_placebo <- lm(murder ~ treat + after_placebo + treatafter_placebo, data = DiD_placebo)

tidy(mod_placebo) %>%
  filter(term == "treatafter_placebo") %>%
  transmute(
    term = "Placebo DiD effect (pretend treatment starts in 2012)",
    estimate,
    std.error,
    p.value,
    interpretation = case_when(
      p.value < 0.05 & estimate > 0 ~ "Evidence against no anticipation: treated outcomes rise before the true treatment year",
      p.value < 0.05 & estimate < 0 ~ "Evidence against no anticipation: treated outcomes fall before the true treatment year",
      TRUE ~ "No strong evidence of anticipation effects"
    )
  ) %>%
  kable(digits = 3)
```

> **Exercise 2:** If the placebo coefficient is large, what does that suggest about the no-anticipation assumption? What if it is close to zero?

# 5. Then evaluate parallel trends

## What parallel trends means

Parallel trends means that, without legalization, treated and control states would have experienced the same changes in murder rates between the pre and post periods.

## Discussion questions

- What evidence would make you more comfortable with the parallel trends assumption?
- What evidence would make you doubt it?
- How is this dataset limited for testing the assumption?

## Visualizing the trends

```{r parallel-trends-plot}
DiD %>%
  mutate(group = if_else(treat == 1, "Treated", "Control")) %>%
  ggplot(aes(x = year, y = murder, color = group, group = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_vline(xintercept = 2014, linetype = "dashed", color = "gray") +
  labs(
    x = "Year",
    y = "Murder rate",
    color = NULL,
    title = "Murder rate trends for treated and control groups",
    subtitle = "Vertical line marks the 2014 legalization year"
  )
```

## A simple pre-trend check

```{r parallel-trends-pretreatment-plot}
pre_period <- DiD %>%
  filter(year < 2014) %>%
  mutate(
    year_centered = year - min(year),
    group = if_else(treat == 1, "Treated", "Control")
  )

pre_period %>%
  ggplot(aes(x = year, y = murder, color = group, group = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9, linetype = "dotted") +
  labs(
    x = "Year",
    y = "Murder rate",
    color = NULL,
    title = "Pre-2014 murder trends",
    subtitle = "Dotted lines are linear fits estimated only on the pre-treatment period"
  )
```

```{r parallel-trends-test}
pre_period <- DiD %>%
  filter(year < 2014) %>%
  mutate(year_centered = year - min(year))

pretrend_mod <- lm(murder ~ treat * year_centered, data = pre_period)

tidy(pretrend_mod) %>%
  filter(term == "treat:year_centered") %>%
  transmute(
    term = "Differential pre-trend (treated x year)",
    estimate,
    std.error,
    p.value,
    interpretation = if_else(
      p.value < 0.05,
      "Evidence against parallel linear pre-trends",
      "No strong evidence of different linear pre-trends"
    )
  ) %>%
  kable(digits = 3)
```

This does not prove parallel trends. It is only a diagnostic: failing to reject a differential pre-trend is consistent with the assumption, but it does not guarantee that the treated group would have followed the control trend after 2014 in the absence of treatment.

> **Exercise 3:** Based on the pre-2014 plot and the differential pre-trend estimate, do the treated and control groups look broadly consistent with parallel trends? What caveats should you mention?


## Final discussion

- Why is the treatment indicator constant over time for the treated group?
- What would a 2×2 DiD look like if the outcome were unemployment rather than murder?
- How would you explain the ATT to a policy audience?



